Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-06 main.tex
Updates from june

@@

\paragraph{Interpretation.}
As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\natp_0$ : 
\begin{itemize}
	\item Stochastic Bregman Proximal Point with learning rate $\inv{n_0+n}$,
	\item Stochastic Mirror Sescent with learning rate  $\inv{n_0+n +1}$.
\end{itemize}
In both cases, $f(\natp)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.

@@

\section{Stochastic Bregman Proximal Point}
\subsection{Correspondence with MAP}
For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
\begin{align}
	p(\natp |\cD_n)
	=p(\natp|X_n,\cD_{n-1})
	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
\end{align}
For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
\begin{align}
	- \log p(\natp | \cD_n) 
	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
\end{align}
so we immediately get the recursion
\begin{align}
    \hat \natp_n^\text{MAP}
    &= \argmin_\natp - \log p(\natp | \cD_n) \\
    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
\end{align} 
Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.

@@

\paragraph{Stochastic Setting.}
This proof does not transpose immediately to the stochastic setting.
First off, we do not have a decreasing sequence anymore
\begin{align}
	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
\end{align}
Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
The most straightforward such assumption is 
\begin{align}
\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Then, using convexity of $f$, we get the rate
\begin{align}
	 \expect{f(\frac{\sum_{t=1}^{T} \lr_t \natp_{t+1}}{\sum_{t=1}^{T} \lr_t } )} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
In addition the convergence bears on average iterates, which is far from perfect.

@@

\section{Stochastic Mirror Descent (SMD)}
\subsection{MAP as SMD}
Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpart(\natp)$. In words, $f$ is linear modification of a convex function $\logpart$, which we can access only through noisy estimates of $\mu_*$. It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$. First let's recall mirror descent iteration
\begin{align}
	\natp_{t+1} 
	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
\end{align}
where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$. Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$. 
Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
\begin{align}
	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
\end{align}
which shows exponential convergence and 1-step convergence when $\lr =1$. Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

\subsection{Relative Smoothness for Mirror Descent}
In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
In our setting this is not always true -- eg gaussians.
However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
\begin{align}
	& \text{$f$ is $L$-smooth relative to $h$} \\
	\iff & Lh - f \ \text{convex} \\
	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
\end{align} 
where the last equivalence holds only when $f$ and $h$ are twice differentiable. In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians. Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
\begin{align}
	& f-\mu h \ \text{ convex} \\
	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
\end{align}

@@

% todo: idee: adapter la preuve de stochastic gradient descent a stochastic mirror descent. Exploiter un parralele entre ces deux preuves pour obtenir une preuve pour stochastic mirror descent dans le cas ou f est relatively smooth w.r.t. A*. ce qu'on veut, c'est une hypothese analogue a expected smoothness (see paper of nico loizou) pour le relatively smooth.
% SGD: General Analysis and Improved Rates gower et al -- section 2 and 3 see equation (9) in particular (expected smoothnesS)
% Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3

@@

\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant. This convergence rate holds only after a given number of samples have been seen. This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.

@@

\subsection{Alternative Research Tracks}
\begin{itemize}
	\item Weakly supervised : missing part of $x_i$ for each $i$.
	\item Structured prediction : graphical model on $X$ or other kind of structure.
\end{itemize}

@@

>>>>

\subsection{Triangle Scaling Exponents}

\TODO{cite accelerated Bregman gradient paper} 
defines the Triangle Scaling Exponent, to characterize an acceleration Bregman methods. The exponent $\gamma$ of a Bregman divergence $\cB$ is defined as the largest coefficient such that $
	\forall x_0, x_1, x_2, \forall \lambda\in[0,1],$
\begin{align}
	\cB((1-\lambda)x_0 + \lambda x_1 ; (1-\lambda)x_0 + \lambda x_2)
	\leq \lambda^\gamma \cB(x_1 ; x_2) \; .
\end{align}
Geometrically, for any triangle, it tells us a shape within which the geodesic triangle is contained (to clarify with a picture.)
It is related to Thales triangle.
We precisely have a Thales triangle with the dual expectation pivot, meaning we can relate the MAP variance term with the MLE expected suboptimality. Writing $\lambda = \frac{n}{n+n_0}$, we observe
\begin{align}
	\MAPm^\text{MAP} &= (1-\lambda)\mu_0 + \lambda \MAPm^\text{MLE} \\
	\mu_n &= (1-\lambda)\mu_0 + \lambda \mu^*
\end{align}
thus
\begin{align}
	\expect{\bregmanconj(\mu_n; \MAPm^\text{MAP})}
	\leq \left (\frac{n}{n+n_0} \right)^\gamma \expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})} \; .
\end{align}
This relationship between the MAP variance and the MLE may be void, as the MLE expected loss may be infinite. But in some settings it might be helpful to characterize the MAP convergence.
For instance, this same paper says that for a gaussian variance, $\gamma=1$.

@@

\paragraph{Interpretation.}
As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\natp_0$ : 
\begin{itemize}
	\item Stochastic Bregman Proximal Point with learning rate $\inv{n_0+n}$,
	\item Stochastic Mirror Descent with learning rate  $\inv{n_0+n +1}$.
\end{itemize}
In both cases, $f(\natp)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.

@@

\paragraph{\TODO{Discussion with Radu}}
PROBLEM. Even in the Euclidean setting, the decreasing step-size analysis with SGD (or SPP) gives a rate $O(1/\log(n))$, which is far from the asymptotic $O(1/n)$.

If we change the step-size, for $1/\sqrt{n}$ for instance, the algorithm changes. It becomes a weighted average of the sufficient statistics. It might be helpful for online learning (changing target), but it is not relevant when the target is fixed.

\section{Stochastic Bregman Proximal Point}
\subsection{Correspondence with MAP}
For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
\begin{align}
	p(\natp |\cD_n)
	=p(\natp|X_n,\cD_{n-1})
	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
\end{align}
For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
\begin{align}
	- \log p(\natp | \cD_n) 
	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
\end{align}
so we immediately get the recursion
\begin{align}
    \hat \natp_n^\text{MAP}
    &= \argmin_\natp - \log p(\natp | \cD_n) \\
    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
    \label{eq:SBPP}
\end{align} 
Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.

@@

\paragraph{Stochastic Setting.}
This proof does not transpose immediately to the stochastic setting.
First off, we do not have a decreasing sequence anymore
\begin{align}
	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
\end{align}
Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
\begin{align}
	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
	\leq \cB( \natp^*; \natp_{0})  \; .
\end{align}
To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
The most straightforward such assumption is 
\begin{align}
\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
\label{eq:proximal_variance}
\end{align}
e.g. we relate the stochastic function value on the next iterate to the function value.
Then, taking the iterate with the least expected loss, we get the rate
\begin{align}
	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
\end{align}
which is very analogous to standard rates for SGD.
It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.

@@

\section{Stochastic Mirror Descent (SMD)}
\subsection{MAP as SMD}
In this section, we show that MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.

\paragraph{Mirror Descent.}
Let us recall mirror descent iteration
\begin{align}
	\natp_{t+1} 
	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
\end{align}
where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$. Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$. 
Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
\begin{align}
	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
\end{align}
which shows exponential convergence and 1-step convergence when $\lr =1$.

\paragraph{Stochastic Mirror Descent}
Let $f(\natp, x) = -\log p(x|\natp)$.
MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP} and manipulating the Bregman yields
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

@@

@@

\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant. This convergence rate holds only after a given number of samples have been seen. This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum. 
\TODO{Sanity check literature search with Bach's references.}

\TODO{3 threads of literature search: optimization, statistics, learning theory. This last one is not covered yet. It is more interested in finite sample results than statistics. Contact Alexander Rakhlin}

@@

\subsection{Alternative Research Tracks}
\begin{itemize}
	\item A picture / geometric understanding of the equality between ELBO (Jensen) and relative smoothness for EM.
	\item Bregman is about general geometry, can we bring it to the non-convex realm via local analysis ?
	\item Can we design an algorithm with higher order Bregman divergences ? Does it mean anything ? The Bregman is a quadratic approximation of a function, yielding a geometry. Higher order does not carry such meaning. But we would like to involve some quadratic function information within mirror descent !
\end{itemize}

@@

\section{Relative Smoothness for Mirror Descent}
In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
In our setting this is not always true -- eg gaussians.
However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
\begin{align}
	& \text{$f$ is $L$-smooth relative to $h$} \\
	\iff & Lh - f \ \text{convex} \\
	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
\end{align} 
where the last equivalence holds only when $f$ and $h$ are twice differentiable. In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians. Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
\begin{align}
	& f-\mu h \ \text{ convex} \\
	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
\end{align}
